Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-05-07 paper/methodology.tex
last changes before distributing first draft

\begin{figure*}[t!]
\plotone{plot3diagsData.png}
\caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements). A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots. Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations). The same symbol color scheme is used in other two panels. The three yellow lines in the middle panel show stellar locus parametrization used by Green et al. (2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$. In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other. In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars. The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
\end{figure*}

Our approach adopted here is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
to the disk and thus $A_r$ can be obtained from IR maps, they have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I}, can be used to estimate distance
in a straightforward manner.}
by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G},  \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
The main differences compared to these works include: 
\begin{enumerate}
\item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
                blue horizontal branch stars; potentially Miras, quasars and RR Lyrae stars too, which can also be recognized and rejected using variability).
\item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
               and an extended $[Fe/H]$ range. 
\item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
               also account for the Galaxy's bulge component. 
\end{enumerate}

@@

\begin{figure*}[b!]
\plotone{plot3HRdiags.png}
\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
The date are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner.
Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
using eqs. A2 and A7 from Ivezi\'{c} et al. (2008).  Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
Figure~\ref{fig:3dataDiags}, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
\label{fig:3HRdiags}
\end{figure*}

@@

We use a combination of SDSS and Gaia data to demonstrate the first problem. The middle panel in Figure~\ref{fig:3HRdiags} 
shows a clear discrepancy between SDSS-based empirical tracks and data for sub-giant stars, using parallax-based absolute
magnitudes. Nevertheless, it is noteworthy that
for main-sequence stars the agreement is excellent. Furthermore, the right panel in Figure~\ref{fig:3HRdiags} demonstrates
that SDSS-based photometric distances and Gaia-based ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}
for main-sequence stars are on the ``same scale''.  XXX ZI: provide numerical results for offsets!

@@

\begin{figure*}[ht!]
\plotone{RGdegeneracy.png}
\caption{Analysis of color degeneracies between giants and main  sequence stars. The left panel shows minimum color distance in
  4-dimensional color space ($u-g$, $g-r$, $r-i$, $i-z$) between a position on the locus with $M_r<4$ (giants) and any position on the
  main-sequence $M_r>4$ locus (in magnitudes). Symbols are color-coded by metallicity, linearly from blue to red for the range $-2.5$ to 0.5).
  The middle and right panels show errors in absolute magnitude and metallicity when a giant is misidentified as a main sequence star
  closest to it in color space.} 
\label{fig:RGdegeneracy}
\end{figure*}

@@

\begin{figure*}[ht!]
\plotone{plot3diagsBAzoom.png}
\caption{Luminosity-color tracks for H and He white dwarfs based on models from \cite{1995PASP..107.1047B}, with
  colors slightly shifted to bring them in agreement with SDSS photometry. Data, shown as symbols, are the same as in
  Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}. Note the crucial role of the $u-g$ color
  for distinguishing H and He tracks.}
\label{fig:locusWDs}
\end{figure*}

@@

\begin{figure*}[ht!]
\plotone{plot3diagsBA_DAh.png}
\caption{Luminosity-color tracks, shown with gray symbols, for unresolved binaries composed of an M dwarf with
  $[Fe/H]=-1.5$ and an H white dwarf. Analogous tracks for binaries composed of an M dwarf with
  $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical. Data, shown as symbols,
  are the same as in Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}.} 
\label{fig:locusBinaries}
\end{figure*}

@@

\subsection{TRILEGAL-based Priors}

@@

In case of $N_M$ models, we assume that they represent an exhaustive model set and estimate the posterior
probability of model $k$, $k=1...N_M$, as
\begin{equation}
                 p(k) = { O_{k1}  \over \sum_{j=1}^{j=N_M}  O_{kj}},
\end{equation}
where the model $k=1$ is chosen arbitrarily without a loss of generality ($O_{11}$ = 1). We will illustrate model
selection with multiple stellar populations in the next Section.

>>>>

\begin{figure*}[t!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsData.png} 
\caption{An illustration of multiple stellar populations. The dots (of
any color) in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements). A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots. Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations). The same symbol color scheme is used in other two panels. The three yellow lines in the middle panel show stellar locus parametrization used by Green et al. (2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$. In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other. In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars. The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
\end{figure*}

Our approach adopted here is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
to the disk and thus $A_r$ can be obtained from IR maps, they have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I}, can be used to estimate distance
in a straightforward manner.}
by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G}, \cite{lallement_3d_2014}, 
\cite{gordon_panchromatic_2016}, \cite{queiroz_starhorse_2018}, \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
The main differences compared to these works include: 
\begin{enumerate}
\item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
                blue horizontal branch stars; potentially Miras, quasars and RR Lyrae stars too, which can also be recognized and rejected using variability).
\item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
               and an extended $[Fe/H]$ range. 
\item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
               also account for the Galaxy's bulge component. 
\end{enumerate}

@@

\begin{figure*}[b!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3HRdiags.png}  
\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
The data are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner.
Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
using eqs. A2 and A7 from Ivezi\'{c} et al. (2008).  Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
Figure~\ref{fig:3dataDiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
\label{fig:3HRdiags}
\end{figure*}

@@

We use a combination of SDSS and Gaia data to demonstrate the first problem. The middle panel in Figure~\ref{fig:3HRdiags} 
shows a clear discrepancy between SDSS-based empirical tracks and data for sub-giant stars, using parallax-based absolute
magnitudes. Nevertheless, it is noteworthy that
for main-sequence stars the agreement is excellent. Furthermore, the right panel in Figure~\ref{fig:3HRdiags} demonstrates
that SDSS-based photometric distances and Gaia-based ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}
for main-sequence stars are on the ``same scale''.  We discuss these
distance scales in more detail in \S\ref{sec:SDSSGaia}.

@@

\begin{figure*}[ht!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/RGdegeneracy.png}
\caption{Analysis of color degeneracies between giants and main  sequence stars. The left panel shows minimum color distance in
  4-dimensional color space ($u-g$, $g-r$, $r-i$, $i-z$) between a position on the locus with $M_r<4$ (giants) and any position on the
  main-sequence $M_r>4$ locus (in magnitudes). Symbols are color-coded by metallicity, linearly from blue to red for the range $-2.5$ to 0.5).
  The middle and right panels show errors in absolute magnitude and metallicity when a giant is misidentified as a main sequence star
  closest to it in color space.} 
\label{fig:RGdegeneracy}
\end{figure*}

@@

\begin{figure*}[ht!] 
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsBAzoom.png}
\caption{Luminosity-color tracks for H and He white dwarfs based on models from \cite{1995PASP..107.1047B}, with
  colors slightly shifted to bring them in agreement with SDSS photometry. Data, shown as symbols, are the same as in
  Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}. Note the crucial role of the $u-g$ color
  for distinguishing H and He tracks.}
\label{fig:locusWDs}
\end{figure*}

@@

\begin{figure*}[ht!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsBA_DAh.png}  
\caption{Luminosity-color tracks, shown with gray symbols, for unresolved binaries composed of an M dwarf with
  $[Fe/H]=-1.5$ and an H white dwarf. Analogous tracks for binaries composed of an M dwarf with
  $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical. Data, shown as colored symbols,
  are the same as in Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}.} 
\label{fig:locusBinaries}
\end{figure*}

@@

\subsection{TRILEGAL-based Priors \label{sec:priors}}

@@

In case of $N_M$ models, we assume that they represent an exhaustive model set and estimate the posterior
probability of model $k$, $k=1...N_M$, as
\begin{equation}
                 p(k) = { O_{k1}  \over \sum_{j=1}^{j=N_M}  O_{kj}},
\end{equation}
where the model $k=1$ is chosen arbitrarily without a loss of generality ($O_{11}$ = 1).
% We will illustrate model selection with multiple stellar populations in the next Section.

### 2024-05-07 paper/paper.tex
new sections 2.2.3 and 2.2.4, and comments from Lovro and Bob
progress...
last changes before distributing first draft

\usepackage{amssymb,amsmath}
% \usepackage{lsst}
\usepackage{xspace}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{booktabs}
\addto\extrasenglish{%
  \renewcommand{\equationautorefname}{eq.}%
}

@@

% Croatian letter "dj"
\def\d   {{d $\mkern-14.3mu \mathchar'26 $}}

@@

\author[0000-0012-3245-123X]{\v{Z}eljko Ivezi\'{c}}
\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}

@@

@@

%\include{analysis}

%\include{discussion}

\begin{acknowledgements}

This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme.

\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality.

@@

This work has made use of data from the European Space Agency (ESA)
mission Gaia \url{https://www.cosmos.esa.int/gaia}, processed by the Gaia
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/ dpac/consortium}). Funding for the
DPAC has been provided by national institutions, in particular, the
institutions participating in the Gaia Multilateral Agreement.

\end{acknowledgements}

@@

>>>>

\usepackage{amssymb,amsmath}
% \usepackage{lsst}
\usepackage{xspace}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{gensymb}

\addto\extrasenglish{%
  \renewcommand{\equationautorefname}{eq.}%
}

@@

% Croatian letter "dj"
\def\dj   {{d $\mkern-14.3mu \mathchar'26 $}}

@@

\author[0000-0001-5250-2633]{\v{Z}eljko Ivezi\'{c}}
\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}

@@

\author[0000-0000-0000-0000]{Neven Caplar}
\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}

@@

\include{analysis}

\include{pipeline}

\include{discussion}

\begin{acknowledgements}
This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale
de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme. 
\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality. 
B.A. acknowledges helpful discussions with Douglas Tucker and the members of the Rubin Observatory Stack Club.

@@

This work has made use of data from the European Space Agency (ESA)
mission Gaia \url{https://www.cosmos.esa.int/gaia}, processed by the Gaia
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/ dpac/consortium}). Funding for the
DPAC has been provided by national institutions, in particular, the
institutions participating in the Gaia Multilateral Agreement.
\end{acknowledgements}

@@

dust extinction plot in: 
BayesMethod3D_SEGUElong.ipynb

BDBS:
BDBS/analyzeBDBS_field_044_TRILEGALcatalog.ipynb
PhotoD/notebooks/BayesMethod3D_BDBSfield044.ipynb

IBJ2012: fig. 23 
star with g-i = 3.0 and r =20: DM = 7 mag so d=250 pc
Mr = 13

10% distance from both Gaia and LSST: fig. 21: 
sigmaPi = 0.5 mas for both LSST and Gaia
SNR=10, so pi = 5 mas or D=200 pc OK

within 150 pc: 10% dist for DM=6 and r = 21.5 => Mr = 15.5  (g-i = 3.6) 
LSST sigmaPi at r=21.5: 0.7 mas so pi = 7 mas or D = 150 pc!

blue star with g-i = 0.3 and r = 20: DM = 16, d=15 kpc
so Mr = 4
parallax = 1/15 mas so SNR=0.13

DM = 12: r = 16, d=2.5 kpc, pi = 0.4 
Gaia sigmaPi = 0.05 mas, SNR = 10 
LSST pi limited by systematics: sigmaPi = 0.5 so SNR~1

LSST at r = 24: sigmaPi = 3
for SNR=10, pi = 30 mas and d=30pc DM=2.5 mag
Mr = 24-2.5 -> Mr = 21.5

For stars with g-i > 3.2 (M5 and later), LSST has better parallax distance
estimates for r>20

For stars of any color at distances beyond 3 kpc, LSST photo distances
are better than Gaia distances
For blue stars at 3 kpc, r = 16 and Gaia's parallax are already 10\%
between 16 and 20, LSST benefits from FeH which photogeo don't have
LSST at blue end goes to 100 kpc with 10% distance, correct?
